# R refresher {#sec:refresher}

The objectives of this chapter is to review some R syntax, functions
and data structures that will be needed for the following chapters.

## Administration

- Setting up an [RStudio project](https://uclouvain-cbio.github.io/WSBIM1207/sec-startr.html#getting-set-up)
- [Install](https://uclouvain-cbio.github.io/WSBIM1207/sec-startr.html#r-packages)
  packages from CRAN and
  [Bioconductor](https://uclouvain-cbio.github.io/WSBIM1207/sec-bioinfo.html#sec:bioconductor).

```{r, eval = FALSE}
BiocManager::install("UCLouvain-CBIO/rWSBIM1322")
```

- [Avoid saving and loading your
  workspace](https://uclouvain-cbio.github.io/WSBIM1207/sec-startr.html#getting-set-up)
  (the `.RData` file).
- [UTF-8 character encoding](https://uclouvain-cbio.github.io/WSBIM1207/sec-startr.html#getting-set-up).
- Starting a markdown document.

## Basic data structures and operations

- [vectors](https://uclouvain-cbio.github.io/WSBIM1207/sec-startr.html#introduction-to-r),
  generating and subsetting vectors.
- Missing values.
- [Factors](https://uclouvain-cbio.github.io/WSBIM1207/sec-startdata.html#factors)
- [Dataframes](https://uclouvain-cbio.github.io/WSBIM1207/sec-startr.html#introduction-to-r) (and tibbles)
- [Matrices](https://uclouvain-cbio.github.io/WSBIM1207/sec-startdata.html#matrices)
- Arrays
- [Lists](https://uclouvain-cbio.github.io/WSBIM1207/sec-startdata.html#lists)

**Summary**


|            | number of dimensions | number of data types |
|------------+----------------------+----------------------|
| **vector** | 1 (length)           | 1                    |
| **matrix** | 2                    | 1                    |
| **array**  | n                    | 1                    |
| **dataframe** | 2                 | n                    |
| **list**   | 1 (length)           | n                    |


## Tidyvers

- The [dplyr](https://uclouvain-cbio.github.io/WSBIM1207/sec-dplyr.html) package
- [Piping](https://uclouvain-cbio.github.io/WSBIM1207/sec-dplyr.html#pipes)
- Wide and long data (converting with `gather` and `spread` - see also
  the [new `pivot_longer` and `pivot_wider`
  functions](https://tidyr.tidyverse.org/dev/articles/pivot.html))

## Saving and exporting

- `save()` and `load()` binary data.
- `saveRDS` and `readRDS` binary data.
- [Exporting
  data](https://uclouvain-cbio.github.io/WSBIM1207/sec-dplyr.html#exporting-data-1)
  with `write.csv` and `read.csv` (or `write_csv` and `read_csv`) and
  same for other types of spreadsheets.
- [Saving
  figures](https://uclouvain-cbio.github.io/WSBIM1207/sec-vis.html#exporting-plots)
  (`ggsave` and file devices such as `png()`, `pdf()`, ...).
- Provenance: `sessionInfo()`

## Programming

- [Writing
  functions](https://uclouvain-cbio.github.io/WSBIM1207/sec-prog.html#writing-new-functions)
- [Conditionals](https://uclouvain-cbio.github.io/WSBIM1207/sec-prog.html#conditionals)
  `if`/`else`
- [Iteration](https://uclouvain-cbio.github.io/WSBIM1207/sec-prog.html#iteration):
  `for` loops and `apply` functions

## Additional exercises



`r msmbstyle::question_begin()`
Read the `interro2.rds` from the `rWSBIM1207` package (version 0.1.9
of later) file into R. The path to the file can be found with the
`rWSBIM1207::interro2.rds()` function.

This dataframe provides the scores for 4 tests for 10 students.

- Write a function that calculates the average score for the 3 best
  tests.
- Calculate this average score for the 10 students.

This can be done using the `apply` function or using `dplyr`
functions. For the latter, see `rowwise()`

`r msmbstyle::question_end()`


```{r, include = FALSE}
stopifnot(packageVersion("rWSBIM1207") >= "0.1.9")

moy <- function(x) {
    x <- sort(x, decreasing = TRUE)[1:3] ## gets rid of NAs
    mean(x, na.rm = TRUE)
}


interro2 <- readRDS(rWSBIM1207::interro2.rds())
interro2$m <- apply(interro2[, -1], 1, moy)


interro2 <- readRDS(rWSBIM1207::interro2.rds())

library("tidyverse")

interro2 %>%
    rowwise() %>%
    mutate(m = moy(c(interro1, interro2, interro3, interro4)))

interro2 %>%
    gather(key = interro, value = score, -noma) %>%
    group_by(noma) %>%
    summarise(m = moy(score)) %>%
    full_join(interro2)
```
