# Data visualisation {#sec:vis}

This chapter is based on Chapter 3 [*High Quality Graphics in
R*](https://www-huber.embl.de/msmb/Chap-Graphics.html) of the *Modern
Statistics for Modern Biology* book book by Susan Holmes and Wolfgang
Huber.

```{r env, message = FALSE, echo = FALSE}
suppressPackageStartupMessages(library("ggplot2"))
suppressPackageStartupMessages(library("dplyr"))
suppressPackageStartupMessages(library("Hmisc"))
suppressPackageStartupMessages(library("plotly"))
```

## Introduction

The **goal** of data visualisation is to

- interactively explore and understand data, i.e. *exploratory data
  analysis*;
- to present and *communicate results*, whether as a preliminary
  analysis or final results.

Our **learning objectives** are

- Familiarise ourselves with plotting in R using base graphics and
  `ggplot2`;
- Understand the concepts behind the gramar of graphics and `ggplot2`;
- Learn how to visualiste data along 1, 2 or more dimensions;
- Learn about `plotly`, one package for ploducing interactive
  visualisations.

The different between this chapter, and the [*Data
Visualisation*](ttps://uclouvain-cbio.github.io/WSBIM1207/sec-vis.html)
of the introductory WSBIM1207 course is that here we focus on
understanding and communicating data through visualisation, rather
than on the syntax of the plotting packages. Hence, as a introduction,
let's observe a couple of [*visualisation horror picture show*](./horror.pdf) from a
lecture by Wolfgang Huber.

## Base plotting

The default graphics system that comes with R, often called base R
graphics is simple and fast. It is based on the painter's model or
canvas, where different output are directly overlaid on top of each
other.

Below, we display the relation between the optical density of the
deoxyribonuclease (DNase) protein as measure by an enzyme-linked
immunosorbent assay (ELISA) assay for all observations.

```{r dnase, fig.cap = "The default base `plot` function on the `DNase` data."}
head(DNase)
plot(DNase$con, DNase$density)
```

We can add some features on the plot, such vertical dotted lines for
all observed observations and customise the look and feel of the plot
by setting specific arguments to the `plot` function.

```{r dnase2, fig.cap = "Customising a base figure using function arguments and overlaying new graphical features."}
plot(DNase$con, DNase$density,
     xlab = "DNase concentration (ng/ml)",
     ylab = "Optical density",
     pch = 1,
     col = "steelblue")
abline(v = unique(DNase$conc), lty = "dotted")
```

If we wanted to change anything to that figures, we would need to
repeat all the commands and modify accordingly. Any additinal command
would be added to the existing canvas.

`r msmbstyle::question_begin()`
How would you produce a figure that differentiates the different runs
using base graphics?
`r msmbstyle::question_end()`

`r msmbstyle::solution_begin()`
```{r colbyrun}
plot(DNase$con, DNase$density, col = DNase$Run)
```
`r msmbstyle::solution_end()`

`r msmbstyle::question_begin()`
Use the `hist` and `boxplot` functions to produce and
histogram of all optical densities and a boxplot of the densities
split by run.
`r msmbstyle::question_end()`

`r msmbstyle::solution_begin()`
```{r hist_boxplot, fig.width = 14, fig.height = 7, fig.cap = "The base `hist` and `boxplot` functions, the latter using the formula interface."}
par(mfrow = c(1, 2))
hist(DNase$density)
boxplot(density ~ Run, DNase)
```
`r msmbstyle::solution_end()`

The base graphics function are very effective to quickly produce out
of the box figures. However, there is not global overview and
parametrisation of the visualisation. The layout decisions have to be
made up upfront (and if not adequate, the figure needs to be redrawn)
and every aspect of the figure is customised locally as function
arguments.

More generally, base graphics functions will work with various inputs:
above we have worked with a `data.frame`, vectors and a formula. There
is no unified type of data across all functions which makes it
efficient for some types of data (if they match), but also very
heterogeneous in terms of interface, leading to a lot of customisation
code.

Finally, defaults, and colours in particular, are poorly chosen.

## The `ggplot2` package

**`ggplot2`** is a plotting package that makes it simple to create
complex plots from data in a data frame. It provides a more
programmatic interface for specifying what variables to plot, how they
are displayed, and general visual properties. The theoretical
foundation that supports the `ggplot2` is the *Grammar of
Graphics*[^gg]. Instead of producing the figure, the user defines and
assembles the visual components into an object that is the
displayed. There is a book about ggplot2[^ggbook] that provides a good
overview, but it is outdated. The ggplot2 web page
(https://ggplot2.tidyverse.org) provides ample documentation.


[^gg]: Wilkinson, Leland. 2005. The Grammar of Graphics (Statistics
    and Computing). Berlin, Heidelberg: Springer-Verlag.

[^ggbook]: Wickham, Hadley. 2016. ggplot2: Elegant Graphics for Data
    Analysis. Springer-Verlag New York.


To build a ggplot, we will use the following basic template that can
be used for different types of plots:

```
ggplot(data = <DATA>, mapping = aes(<MAPPINGS>)) +  <GEOM_FUNCTION>()
```

We need first to load the `ggplot2` package:

```{r ggload}
library("ggplot2")
```

- use the `ggplot()` function and bind the plot to a specific data frame using the
      `data` argument

```{r gg0, fig.cap = "We have only specified the data, and there's nothing to display yet."}
ggplot(data = DNase)
```

- define a mapping (using the aesthetic (`aes`) function), by
  selecting the variables to be plotted and specifying how to present
  them in the graph, e.g. as x/y positions or characteristics such as
  size, shape, colour, etc.

```{r gg1, fig.cap = "`ggplot2` can now generate the axes, ticks and ranges based on the data."}
ggplot(data = DNase,
       mapping = aes(x = conc, y = density))
```

- add *geoms* – graphical representations of the data in the plot
  (points, lines, bars). To add a geom to the plot use the `+`
  operator. Because we have two continuous variables, let's use
  `geom_point()` first:

```{r gg2, fig.cap = "Final figures with rendering of the data as a scatter plot."}
ggplot(data = DNase,
       mapping = aes(x = conc, y = density)) +
  geom_point()
```


`r msmbstyle::question_begin()`
Compare the `ggplot2` and base graphics version of the density
vs. concentration plot. Which one do you prefer, and why?
`r msmbstyle::question_end()`

It is possible to store the output of the `ggplot` function into a
variable that can be visualised by either typing its name in the
console or explicitly printing it (like any other variable).

```{r ggobj, fig.cap = "Saving and printing an object."}
gg <- ggplot(data = DNase,
             mapping = aes(x = conc, y = density)) +
    geom_point()
print(gg)
```

Let's immediately customise this visualisation to
- highlight how to re-use the `gg` object without repeating the
  plotting code and
- how we can add additional (identical or different) *geoms* to a
  plot.

```{r gg3, fig.cap = "Adding another `geom_point` with its own (local) aesthetics. "}
gg + geom_point(aes(colour = Run))
```


`r msmbstyle::question_begin()`
What do you think of the colours used to differentiate the different
runs above?
`r msmbstyle::question_end()`

`r msmbstyle::solution_begin()`
```{r ggex, fig.cap = "Setting the Run to a category by coercing it to a character, rather than a numeric."}
gg + geom_point(aes(colour = as.character(Run)))
```
`r msmbstyle::solution_end()`

Below is an example of more elaborated composition, overlaying points
and a non-linear *loess* regression. But first, let's load a object
containing microarray data from the `hiiragi2013df1` data. The data is
originally available from the
[`Hiiragi2013`](http://bioconductor.org/packages/release/data/experiment/html/Hiiragi2013.html)
package and describes the cell-to-cell expression variability followed
by signal reinforcement progressively segregates early mouse lineages
[@Ohnishi:2014].


```{r gg4, fig.cap = "Modelling the relation between the expression of X1426642_at and X1418765_at."}
library(rWSBIM1322)
data(hiiragi2013df1)
ggplot(hiiragi2013df1, aes(x = X1426642_at, y = X1418765_at)) +
    geom_point(shape = 1) +
    geom_smooth(method = "loess")
```

And, adding colours representing the different samples

```{r gg5, fig.cap = "Modelling the relation between the expression of X1426642_at and X1418765_at and annotating samples."}
ggplot(hiiragi2013df1, aes(x = X1426642_at, y = X1418765_at))  +
  geom_point(aes(color = sampleColour), shape = 19) +
  geom_smooth(method = "loess")
```

## Visualising 1D data

Let's start by exploring some 1 dimensional visualisation. This is
very relevant for omics data such as transcriptomics or quantitative
proteomics, when contrasting the expression values across multiple
samples.


First, we convert a microarray gene expression data to a data.frame,
fit for some `ggplot2` visualisation, focusing on genes Fgf4, Gata4,
Gata6 and Sox2.


```{r selprobes}
selected_genes <- c(Fgf4 = "X1420085_at", Gata4 = "X1418863_at",
                    Gata6 = "X1425463_at",  Sox2 = "X1416967_at")
dfx <- hiiragi2013df1[, selected_genes]
```

```{r gather}
library("dplyr")
library("tidyr")
genes <- gather(dfx, key = "gene", value = "expression")
head(genes)
```

### Histogram

```{r hist, fig.cap = "Distribution of the Gata4 expression"}
genes %>%
    filter(gene == "Gata4") %>%
    ggplot(aes(x = expression)) + geom_histogram()
```


### Box plots

```{r boxplot, fig.cap = "A boxplot of expression values."}
p <- ggplot(genes, aes(x = gene, y = expression, fill = gene))
bxplot <- p + geom_boxplot()
bxplot
```


`r msmbstyle::question_begin()`
Repeat the above figure replacing the boxes by violins using the
`geom_violin`. Which one do you think does a better job?
`r msmbstyle::question_end()`


`r msmbstyle::solution_begin()`
```{r violin, fig.cap = "A violin plot of expression values."}
vlnplot <- p + geom_violin()
vlnplot
```
`r msmbstyle::solution_end()`


### Dot plots and beeswarm plots

When the data aren't too large, it is also possibly to visualise all
points to get a sense of their distribution.


```{r jitter}
jtrplot <- p +
    geom_jitter(aes(colour = gene)) +
    theme(legend.position = "none")
```

In a dotplot, the position of the points along the y axis is
discretised into bins (set as 1/6 below) and the points are then
stacked next to each other.

```{r dot}
dotplot <- p + geom_dotplot(binaxis = "y", binwidth = 1/6,
                            stackdir = "center", stackratio = 0.75,
                            aes(color = gene)) +
    theme(legend.position = "none")
```

The beeswarm algorithms tries to avoid overlapping points.

```{r bee}
library("ggbeeswarm")
beeplot <- p + geom_beeswarm(aes(color = gene)) +
    theme(legend.position = "none")
```

```{r showdots, fig.cap = "Showing all expression values using jittering (left), a dotplot (centre) and a beeswarn plot.", fig.with = 16, fig.height = 4}
library(patchwork)
jtrplot + dotplot + beeplot
```

### Density and ecdf plots


```{r dens}
densplot <- ggplot(genes, aes(x = expression, color = gene)) +
    geom_density() +
    theme(legend.position = "none")
```

```{r ecdf}
ecdfplot <- ggplot(genes, aes(x = expression, color = gene)) +
    stat_ecdf() +
    theme(legend.position = "none")
```


```{r, fig.cap = "Density and cumulative density functions of expression values.", fig.with = 12, fig.height = 6}
densplot + ecdfplot
```

### Summary

- **Boxplot** makes sense for unimodal distributions (see below).

- **Histogram** requires definition of bins (width, positions) and can
  create visual artefacts especially if the number of data points is
  not large.

- **Density** requires the choice of bandwidth; obscures the sample
  size (i.e. the uncertainty of the estimate).

- **ecdf** does not have these problems; but is more abstract and
  interpretation requires more training. Good for reading off
  quantiles and shifts in location in comparative plots.

- **beeswarm**: for up to a few dozens of points, just show the data.


## Number of modes

The number of modes of a distribution depends on scale transformation
of the data.

```{r sim, fig.cap = "Histograms of the same data without (left) and with (right) log-transformation."}
sim <- data.frame(x = exp(rnorm(n = 1e5,
                                mean = sample(c(2, 5),
                                              size = 1e5,
                                              replace = TRUE))))

p1 <-  ggplot(sim, aes(x)) +
    geom_histogram(binwidth = 10, boundary = 0) +
    xlim(0, 400)
p2 <-  ggplot(sim, aes(log(x))) +
    geom_histogram(bins = 30)
p1 + p2
```

This also applies to density plots.

## Visualising 2D data

```{r sc1, fig.cap = "Scatter plot comparing the expression of a wild-type vs. FGF4 KO."}
data(hiiragi2013df2)
head(hiiragi2013df2)
scp <- ggplot(hiiragi2013df2,
              aes(x= `59 E4.5 (PE)`,
                  y = `92 E4.5 (FGF4-KO)`))
scp + geom_point()
```

`r msmbstyle::question_begin()`
The over-plotting of the dots stops us from learning anything about
the density of the different regions of the plot. Use the `alpha`
parameter to `geom_points` between 0 (full transparency) to 1 (opaque,
default).
`r msmbstyle::question_end()`

`r msmbstyle::solution_begin()`
```{r scp2, fig.cap = "Transparency can help distinguish regions of different densities."}
scp + geom_point(alpha = 0.1)
```
`r msmbstyle::solution_end()`


```{r scp3, fig.cap = "Focusing on contours rather that individual values."}
scp + geom_density2d(h = 0.5, bins = 60)
```

```{r scp5, fig.cap = "Local density summaries."}
scp + geom_hex()
```

## Visualising data along more dimensions

When visualising data along additional dimension, we can parameterise
the points by setting their shape, colour, size and transparency, that
can be set with point aesthetics such as `fill`, `color` (or
`colour`), `shape`, `size` and `alpha`.

A very powerful way to represent data along additional dimensions is
facetting, i.e. producing sub-plots for different subsets of the
data. Below, we first re-annotate the data using some regular
expressions


```{r facet1, fig.cap = "Different sub-plots for different lineages using colours (left) of facets(right) to distinguish the different lineages."}
p1 <- ggplot(hiiragi2013df1, aes(x = X1426642_at, y = X1418765_at, colour = lineage)) +
  geom_point()
p2 <- ggplot(hiiragi2013df1, aes(x = X1426642_at, y = X1418765_at)) +
  geom_point() +
    facet_grid( . ~ lineage )
p1 + p2
```

```{r facet2, fig.cap = "Different sub-plots for different lineages and embryonic stages."}
ggplot(hiiragi2013df1,
       aes(x = X1426642_at, y = X1418765_at)) +
    geom_point() +
    facet_grid( Embryonic.day ~ lineage )
```

`r msmbstyle::question_begin()`
Use facets to visualise the distribution of the four Fgf4, Gata4,
Gata6 and Sox2 genes in the `genes` data using histograms.
`r msmbstyle::question_end()`


`r msmbstyle::solution_begin()`
```{r hist2}
ggplot(genes, aes(x = expression)) +
    geom_histogram() +
    facet_wrap(~ gene)
```
`r msmbstyle::solution_end()`


## Interactive visualisation

```{r plotly, fig.cap = "Interactive visualisation with plotly"}
library("plotly")

scp <- ggplot(hiiragi2013df2[1:100, ],
              aes(x= `59 E4.5 (PE)`, y = `92 E4.5 (FGF4-KO)`))

scp2 <- scp + geom_point()
ggplotly(scp2)
```

See https://plot.ly/r for  examples of interactive graphics online.


## Additional exercise

`r msmbstyle::question_begin()`

Load the `barley` data that comes with the `lattice` package (that is
already installed on your computer). The data describe the yield from
a Minnesota barley trial as a data frame with 120 observations on the
following 4 variables.

- `yield` Yield (averaged across three blocks) in bushels/acre.

- `variety` Factor with levels `Svansota`, `No. 462`, `Manchuria`,
  `No. 475`, `Velvet`, `Peatland`, `Glabron`, `No. 457`,`Wisconsin No. 38`, `Trebi`.

- `year` Factor with levels `1932` and `1931`.

- `site` Factor with 6 levels: `Grand Rapids`, `Duluth`,
  `University Farm`, `Morris`, `Crookston`, `Waseca`.

Familiarise yourself with the data
`r msmbstyle::question_end()`


```{r bar1, echo = FALSE}
library(lattice)
data(barley)
```

`r msmbstyle::question_begin()`

Visualise the yield for each variety on all site in 1932 and 1931
using `ggplot2`. Try to generate a figure on your own. If you
struggle, discuss with your fellow students. If you need additional
help, or as a check, execute the code in the `barley` manual page
example.

`r msmbstyle::question_end()`

```{r, echo = FALSE, include = FALSE}
ggplot(barley, aes(x = yield, y = variety, colour = year)) +
    geom_point() +
    facet_grid(site ~ .)
```

`r msmbstyle::question_begin()`
Can you spot an issue with the data?
`r msmbstyle::question_end()`
