# Manipulating sequences with Biostrings {#sec:biostrings}

**Learning objectives**

In this chapter, we will learn about a new object, that is
specialised for the manipulation of biological sequences.

- Learn about the fasta format format for biological sequences.
- Learn about the `*StingSet` classes.
- Working with whole genomes.
- Find patterns in sequence data.

This chapter is based on the course material presented by Martin
Morgan during the [Bioinformatics Summer School
2019](https://uclouvain-cbio.github.io/BSS2019/) at Louvain-la-Neuve,
Belgium.

## Working with sequences - a first start

It is easy to use standard R character vectors to define DNA
sequences. Below, we create a vector with three short sequences.


```{r seq1}
sequences <- c("AAATCGA", "ATACAACAT", "TTGCCA")
sequences
```

We can ask about properties of these sequences and perform some
operations using base R function:


```{r seqop}
length(sequences)
nchar(sequences)
sequences[c(1, 3)]
sample(sequences)
```

However, once is we want to perform biologically relevant operations,
base R fails us; it has no notion of operations relevant to DNA
sequences, e.g.,

```{r, eval=FALSE}
reverseComplement(sequences)
```

Likewise, we can name a variable anything, the semantic meaning
of the variable name is not enforced by R


```{r}
not_a_dna_sequence <- "I love R"
```

## Working with sequences - using `Biostrings`

The Bioconductor `r BiocStyle::Biocpkg("Biostrings")` 
package can do all that base R can do, in addition to knowing about
the semantics of the sequences is handles. Let's start by loading the
package:

```{r loadbiostr, message = FALSE}
library("Biostrings")
```

### `DNAStringSet()`

The `DNAStringSet` class that *knows* about DNA sequences. We can
easily create a `DNAStringSet` from our character vector with

```{r dna1}
dna <- DNAStringSet(sequences)
dna
```

`r msmbstyle::question_begin()`
Does the object `dna` support the operations illustrated above for a
character vector, especially `length()`, `nchar()`, `[`, and
`sample()`?
`r msmbstyle::question_end()`

`r msmbstyle::solution_begin()`
```{r dnaex1}
length(dna)
nchar(dna)
dna[c(1, 3)]
sample(dna)
```
`r msmbstyle::solution_end()`


`r msmbstyle::question_begin()`
Prove to yourself that at least some other useful, DNA-specific,
functions exist, e.g., `reverse()` and `reverseComplement()`.
`r msmbstyle::question_end()`

`r msmbstyle::solution_begin()`

```{r dnaex2}
reverse(dna)
reverseComplement(dna)
```
`r msmbstyle::solution_end()`


`r msmbstyle::question_begin()`
What happens when you try to create a `DNAStringSet()` from an object
such as `not_a_dna_sequence`, defined above, that does not contain a
DNA sequence? Warning: the message is quite cryptic, can you provide a
'human' translation?

`r msmbstyle::question_end()`

`r msmbstyle::solution_begin()`
```{r dnaex3}
try(DNAStringSet(not_a_dna_sequence))
```
`r msmbstyle::solution_end()`


`r msmbstyle::question_begin()`
Why does `DNAStringSet("ACGTMRW")` not create an error, since `MRW`
are not standard nucleotides? For hints, see the section 'The DNA
alphabet:" on the help page `?DNAString`.
`r msmbstyle::question_end()`


`r msmbstyle::question_begin()`
What is the difference between a `DNAString` and a `DNAStringSet`?
`r msmbstyle::question_end()`

`r msmbstyle::solution_begin()`
```{r}
DNAString("AGACCATG")
DNAStringSet(c("AGACCATG", "AAAAAATTTA"))
```
`r msmbstyle::solution_end()`


### Learning more about a class

The function `DNAStringSet()` returns an _object_ that has a
particular class

```{r}
class(dna)
```

Associated with the class are a series of _methods_ (i.e. functions
that have a behaviour that is tuned for that class) that operate on
the class.


`r msmbstyle::question_begin()`
Discover many (unfortunately, not all) methods acting on
an object of class `DNAStringSet` using `methods(class =
"DNAStringSet")`. Verify that `reverseComplement` is among those
methods.
`r msmbstyle::question_end()`

Help pages describing a particular method can be found using `?`, with
the search query quoted and with tab-completion providing hints on
what the appropriate help topic is.


`r msmbstyle::question_begin()`
Find the help page for the `reverseComplement` method operating on a
`DNAStringSet` object, using `?reverseComplement`.
`r msmbstyle::question_end()`


Help pages provide a description of the technical details required for
creating classes and using methods. Vignettes provide a more narrative
description of overall package use.

`r msmbstyle::question_begin()`
Use `browseVignettes(package = "Biostrings")` to see vignettes
available for this package; explore a few vignettes to get a sense of
possible content.
`r msmbstyle::question_end()`

## Reading DNA sequence data from a file

It is unlikely that we would enter 1000's of DNA sequences 'by
hand'. Instead, we might read the data from a standard file
format. For DNA sequences the standard file format is often a 'FASTA'
file, sometimes abbreviated with an extension `.fa` and often
compressed with an additional extension `.fa.gz`. An example of a
FASTA file containing DNA sequences of the 2000bp upstream nucleotides
of all genes annotated in the _Drosophila melanogaster_ `dm3` genome
build, is distributed with the `Biostrings` package. Here's the path
to the FASTA file.

```{r fa_file}
fa_file <-
    system.file(package="Biostrings", "extdata", "dm3_upstream2000.fa.gz")
```


`r msmbstyle::question_begin()`
Take a peak at the structure of a FASTA file by looking at the first
five lines. You can use the `readLines` function, setting the `n`
parameter to read only a limited number of files (see `readLines`).
`r msmbstyle::question_end()`

`r msmbstyle::solution_begin()`
```{r faex1}
readLines(fa_file, 5)
```
`r msmbstyle::solution_end()`

In this case, the first line is an identifier, containing information
about the gene `NM_078863` as well as the genomic coordinates of the
sequence `chr2L:16764737-16766736`. The next lines are the DNA
sequence. After a certain number of lines, a new record starts.

```{r}
tail(readLines(fa_file, 44), 5)
```

We could fairly *easily* write our own parser for this format, but
this would be error-prone and unnecessary. Instread, we want to use
the `readDNAStringSet` function from the `Biostrings` package for
that.

```{r}
dna <- readDNAStringSet(fa_file)
dna
```


`r msmbstyle::question_begin()`
Query the object for basic properties, e.g., it's `length()` and that
the number of characters in each sequence is 2000
`unique(nchar(dna))`.
`r msmbstyle::question_end()`


`r msmbstyle::question_begin()`
Use `letterFrequency()` to determine GC content of each of the DNA
sequences in `dna`. The `letters` argument should be `"GC"`; `as.prob
= TRUE` returns values between 0 and 1. The data is returned as a
matrix with 1 column.
`r msmbstyle::question_end()`

`r msmbstyle::solution_begin()`
```{r letterfreqex1}
gc_content <- letterFrequency(dna, letters = "GC", as.prob = TRUE)
gc_content
```
`r msmbstyle::solution_end()`


`r msmbstyle::question_begin()`
What are the mean and standard deviation of the GC contents in the `r length(dna)` genes?
What genes(s) has the highes GC content?
`r msmbstyle::question_end()`

`r msmbstyle::solution_begin()`
```{r letterfreqex2}
mean(gc_content)
sd(gc_content)
range(gc_content)
dna[which(gc_content == max(gc_content))]
```
`r msmbstyle::solution_end()`


`r msmbstyle::question_begin()`
Plot the distribution of GC frequencies in the `dna` object using base
graphics `hist()` and `plot(density())`, and using `ggplot()`.
`r msmbstyle::question_end()`


`r msmbstyle::solution_begin()`
```{r letterfreqex3a}
hist(gc_content)
plot(density(gc_content))
```

```{r letterfreqex3b}
library("tidyverse") 
p <- as_tibble(gc_content) %>%
    ggplot(aes(x = `G|C`))
p + geom_histogram()
p + geom_density()
```
`r msmbstyle::solution_end()`


## Tidy _Bioconductor_?

Although _Bioconductor_ emphasizes formal objects like `DNAStringSet`
rather than `tibble`-like data frames, some of the ways one interacts
with tidy data can be applied to _Bioconductor_ objects. For instance,
the GC content example might be written in 'traditional' form as

```{r}
gc <- letterFrequency(dna, "GC", as.prob = TRUE)
```

but could be written using pipes and to reesult in a tibble for easier
down-stream manipulation

```{r}
gc <- 
    dna %>%
    letterFrequency("GC", as.prob = TRUE) %>%
    tibble::as_tibble()
gc
```


## Working with genomes

`BSgenome` packages inside Bioconductor contain whole genome sequences
as distributed by ENSEMBL, NCBI and others. In this next example we
will load the whole genome sequence for Homo sapiens from UCSC's hg19
build, and calculate the GC content across chromosome 14.


```{r bs1, message = FALSE}
library("Biostrings")
library("BSgenome.Hsapiens.UCSC.hg19")
chr14_range <- GRanges("chr14", IRanges(1, seqlengths(Hsapiens)["chr14"]))
chr14_range
```

```{r bs2, message = FALSE}
chr14_dna <- getSeq(Hsapiens, chr14_range)
chr14_dna
letterFrequency(chr14_dna, "GC", as.prob=TRUE)
```

`r msmbstyle::question_begin()`
Using `letterFrequency`, calculate the number and frequency of A, G, C
and T nucleotides in chromosome 14.
`r msmbstyle::question_end()`

`r msmbstyle::solution_begin()`
```{r acgtex}
letterFrequency(chr14_dna, c("A", "G", "C", "T"))
letterFrequency(chr14_dna, c("A", "G", "C", "T"), as.prob = FALSE)
```
`r msmbstyle::solution_end()`


## Additional exercises


`r msmbstyle::question_begin()`
Visit the UniProt web page at
[https://www.uniprot.org/](https://www.uniprot.org/) and download the
manually reviewed complete human proteome (containing about 20500
proteins). Do to so, choose *Proteomes* on the main page, then search
for *Homo sapiens*, select it among the list of hits, choose the
reviewed Swiss-Prot release, and download all entries in compressed
fasta format.
`r msmbstyle::question_end()`


`r msmbstyle::question_begin()`
Load the file into R. Beware though that these are protein sequences,
composed of amino acids. Read the `?readDNAStringSet` manual page to
find a more appropriate function.
`r msmbstyle::question_end()`
