# Visualising biomolecular data {#sec:biovis}

The goal of this chapter is to learn some additional visualisation
that are widely used in biomedical data analysis, namely

- **Heatmaps**, including an application of the hierarchical
  clustering that was seen in chapter \@ref(sec:ul).
- Visualisation of intersections, in particular **Venn** and **UpSet**
  plots.

<!-- - The `r BiocStyle::Biocpkg("Gviz")` to visualise genomic data. -->


## Heatmaps


```{r heatmapmulvey, cache = TRUE, fig.cap = "Heatmap of the (normalised) Mulvey et al. 2015 proteomic data."}
data(mulvey2015norm)
heatmap(MSnbase::exprs(mulvey2015norm))
```

A heatmap is composed of two hierarchical clusters (one along the
rows, one along the columns, leading to their re-ordering based on
their similarity) and a intensity matrix. Each of these components is
subject to parameters and options.

As we have seen above, the distance used for clustering can have a
substantial effect on the results, which is conformed below.


```{r heatmapcor, cache = TRUE, fig.cap = "Heatmap of the (normalised) Mulvey et al. 2015 proteomic data using correlation distances."}
heatmap(MSnbase::exprs(mulvey2015norm), distfun = cor.dist)
```


Another important argument, `scale` controls whether rows, columns or
none are scaled. Let's re-use the toy data from the hierarchical
clustering section below.



```{r heatmapscale1, fig.cap = "Effect of scaling on the heatmap output: no scaling."}
library("rWSBIM1322")
data(g3)
heatmap(g3, scale = "none", main = "No scaling")
```

```{r heatmapscale2, fig.cap = "Effect of scaling on the heatmap output: scaling along the rows."}
library("rWSBIM1322")
data(g3)
heatmap(g3, scale = "row", main = "Scaling along rows")
```

```{r heatmapscale3, fig.cap = "Effect of scaling on the heatmap output: scaling along the columns."}
library("rWSBIM1322")
data(g3)
heatmap(g3, scale = "column", main = "Scaling along columns")
```

Based on the caveats above, it is essential to present and interpret
heatmaps with great care.


There exists several packages that allow to produce heatmaps with
various levels of sophistication, such as `heatmap.2` from the `r
CRANpkg("gplots")` package, the `r Biocpkg("Heatplus")` package, or
the `r Biocpkg("ComplexHeatmap")` packages (full documentation
[here](https://jokergoo.github.io/ComplexHeatmap-reference/book/index.html)),
demonstrated below.


```{r complexheatmap, fig.fullwidth = TRUE, message = FALSE, fig.cap = "An annotated heatmap produced with the `ComplexHeatmap` Bioconductor package."}
library("ComplexHeatmap")
x <- MSnbase::exprs(mulvey2015norm)
hcl <- hclust(dist(x))
cl <- cutree(hcl, k = 12)
ha1 <- HeatmapAnnotation(time = mulvey2015norm$time)
ha2 <- HeatmapAnnotation(boxplot = anno_boxplot(x))
ha3 <- rowAnnotation(cluster = factor(cl))
Heatmap(x,
        top_annotation = ha1,
        bottom_annotation = ha2,
        column_names_gp = gpar(fontsize = 8),
        row_names_gp = gpar(fontsize = 3)) +
    ha3
```

Other powerful packages to generate and customise heatmaps are
[superheat](https://rlbarter.github.io/superheat/) and
[pheatmap](https://github.com/raivokolde/pheatmap).


Finally, the `r Githubpkg("talgalili/heatmaply")`,
`r Githubpkg("rstudio/d3heatmap")` and [`iheatmapr`](https://ropensci.github.io/iheatmapr/index.html)
packages can be used to generate interactive heatmaps.


```{r heatmaply, eval = FALSE}
library("heatmaply")
heatmaply(MSnbase::exprs(hlm)[1:100, ])
```

```{r heatmaply2, eval = FALSE}
heatmaply(MSnbase::exprs(hlm)[1:100, ],
          RowSideColors = as.factor(fData(hlm)$markers[1:100]))
```

```{r d3hm, eval = FALSE}
library("d3heatmap")
d3heatmap(MSnbase::exprs(hlm))
```

See also *A tutorial in displaying mass spectrometry-based proteomic
data using heat maps* [@Key:2012], that applies to any type of omics
data (not only proteomics) for a useful reference.


## Visualising intersections between sets

Computing and visualising intersections is a common task in data
analysis. Venn and Euler diagrams are popular representation when
comparing sets and their intersection. Two useful R packages to
generate such plots are
`r CRANpkg("venneuler")` and `r Githubpkg("js229/Vennerable")`.

We will use the `mulvey2015` feature names to generate a test data:

```{r venndat}
set.seed(123)
library("pRolocdata")
data(mulvey2015)
feat_list <- replicate(3,
                       sample(featureNames(mulvey2015), 555),
                       simplify = FALSE)
names(feat_list) <- LETTERS[1:3]
```

The `Venn` function from the `Vennerable` package takes a list as
input, and computes all possible intersections between these elements
of the list. In the output below

- `000` refers to the empty set that are present in none of the
  element of the list;
- `001` is the set of items that are unique to the third element
  (named `C`) of our list;
- ...
- `011` is the set of items that is shared by the second (`B`) and
  third (`C`) element (and absent from the first one) of our list;
- `111` is the set of items that are shared between all elements of
  our list.

```{r venn}
library("Vennerable")
(v <- Venn(feat_list))
```
Each of these intersections can be accessed using through the
`IntersectionSets` slot.

```{r venn2}
## access the identifiers
v@IntersectionSets[["110"]]
```

And finally, the `Venn` object can directly be plotted (albeit with a
suspicious set of colours) with

```{r venn3, fig.cap = "A Venn diagram representing the size of all intersections of the three elements of out `feat_list` input."}
plot(v)
```

Venn diagrams are however limited to two to three, possibly four
sets. The `r CRANpkg("UpSetR")` package is a great solution when
more sets need to be compared.
The `r CRANpkg("UpSetR")` visualises intersections of sets as a matrix
in which the rows represent the sets and the columns represent their
intersection sizes. For each set that is part of a given intersection,
a black filled circle is placed in the corresponding matrix cell. If a
set is not part of the intersection, a light grey circle is shown. A
vertical black line connects the topmost black circle with the
bottom most black circle in each column to emphasise the column-based
relationships. The size of the intersections is shown as a bar chart
placed on top of the matrix so that each column lines up with exactly
one bar. A second bar chart showing the size of the each set is shown
to the left of the matrix.

We will first make use of the `fromList` function to convert our list
to a `UpSetR` compatible input and then generate the figure:

```{r upset1, message = FALSE}
library("UpSetR")
upset_in <- fromList(feat_list)
upset(upset_in)
```

The following
[tweet](https://twitter.com/ngehlenborg/status/986354574989709312) by
the author of the package illustrates how Venn and upset diagrams
relate to each other.


```{r upset2}
upset(upset_in, order.by = "freq")
upset(upset_in, order.by = "degree")
upset(upset_in, order.by = c("freq", "degree"))
upset(upset_in, order.by = c("degree", "freq"))

upset(upset_in, sets = c("A", "B"))
upset(upset_in, sets = c("B", "C", "A"), keep.order = TRUE)

upset(upset_in, group.by = "sets")

## Add set D with a single intersection
upset_in_4 <- upset_in
upset_in_4$D <- 0
upset_in_4[1, "D"] <- 1
head(upset_in_4)

upset(upset_in_4)
upset(upset_in_4, empty.intersections = "on")
```
Visualising intersections with `UpSetR` shines with more that 4 sets,
as Venn diagrams become practically useless.


There is also am UpSetR online app: https://gehlenborglab.shinyapps.io/upsetr/

`r msmbstyle::question_begin()`

Generate a bigger dataset containing 10 sets. Try to generate Venn and
upset diagrams as shown above.

When the number of sets become larger, the options above, as well as
`nsets`, the number of sets (default is 5) and `nintersects`, the
number of intersectios (default is 40) become useful.

`r msmbstyle::question_end()`

`r msmbstyle::solution_begin()`

```{r upsetchallenge}
set.seed(123)
feat_list_10 <- replicate(10, sample(featureNames(mulvey2015), 555), simplify = FALSE)
names(feat_list_10) <- LETTERS[1:10]
feat_list_10 <- fromList(feat_list_10)
upset(feat_list_10)
```

`r msmbstyle::solution_end()`


<!-- ## Data along genomic coordinates -->
