# Conclusions {#sec:ccl}

In this course, we have consolidated our skills in data analysis and
visualisation using R. In particular, this course has focused on
interpretation and understanding of the outputs.

We have also learned and applied new tools, in particular how to
manipulate sequence data using `Biostrings` and how to use statistical
learning tools to explore data and identify patterns of biological
interest. In terms of statistical testing and machine learning, it can
be useful to summarise the different classes of techniques we have
touch on using the figures below. The grid on each of these figures
represents a matrix with quantitative values with features (genes,
proteins, transcripts, ...)  along the rows and samples along the
columns. Annotations of features or samples are presented as coloured
boxes on the right or top if the matrices.


In hypothesis test, we start with quantitative data and an
experimental design, i.e. sample annotation that group samples in
biologically relevant groups. The output of hypothesis testing is a
set of metrics for each features (a p-value, an adjusted p-value, a
fold-change, ...) that informs whether that feature shows any
difference between the biological groups of interest.

```{r, echo = FALSE, fig.cap = "Hypothesis testing.", fig.with = 5, fig.height = 3}
par(oma = c(0, 0, 0, 0), mar = c(0, 0, 2, 0))
rWSBIM1322:::ht_summary()
```

When performing clustering (unsupervised machine learning), we only have our
quantitative data as input, and the clustering algorithm (whether
k-means, hierarchical clustering among many others) suggests a set of
groups to cluster the features (as shown below) or samples.

```{r, echo = FALSE, fig.cap = "Clustering of features.", fig.with = 5, fig.height = 3}
par(oma = c(0, 0, 0, 0), mar = c(0, 0, 2, 0))
rWSBIM1322:::ul_summary()
```

When performing dimensionality reduction with, for example PCA, one
starts with a *n* by *m* data set as input to reduce the number of
dimensions in either direction. On the figure below, the number of
features *n* was reduced to 2 to, typically, visualise the sample
along a scatter plot.

```{r, echo = FALSE, fig.cap = "Dimensionality reduction.", fig.with = 5, fig.height = 3}
par(oma = c(0, 0, 0, 0), mar = c(0, 0, 2, 0))
rWSBIM1322:::dr_summary()
```


In classification (supervised machine learning), we need labelled
data, i.e. a set of sample (top on the figures below) or features
(bottom on the figure below). The classifier uses the data to learn to
assign labels and applies that learnt model it infer the most likely
label of the unlabelled data.

```{r, echo = FALSE, fig.cap = "Classification of samples (top) or features (bottom).", fig.with = 5, fig.height = 6}
par(oma = c(0, 0, 0, 0), mar = c(0, 0, 2, 0))
rWSBIM1322:::sl_summary()
```


The next steps of the curriculum (course
[WSBIM2122](https://uclouvain-cbio.github.io/WSBIM2122/])) will build
upon the skills gained in this course to fully analyse complete
datasets from omics technologies, using state-of-the-art statistical
and machine learning methods and software. The course will be project
based: each experiment and associated technologies will be introduced,
the analysis pipeline will be explained, and the students will then
implement and present the data analysis and critical interpretation of
the results.

## Additional exercise

This final exercise integrates several techniques seen throughout this
course.

`r msmbstyle::question_begin()`
The the RNA-Seq data and sample annotation in the files returned by
the `rWSBIM1207::kem2.tsv()` function. How many genes have been
assayed? Describe the experimental design at hand.
`r msmbstyle::question_end()`


```{r, include = FALSE}
e <- as.matrix(read.delim(rWSBIM1207::kem2.tsv()[1],
                          row.names = 1))
dim(e)
```

`r msmbstyle::question_begin()`
Print and describe the experimental design.
`r msmbstyle::question_end()`

```{r, include = FALSE}
## Below, we harmonise the samples names to have dots in the
## experimental design dataframe (pd: phenotypic data) and expression
## matix.
pd <- readr::read_tsv(rWSBIM1207::kem2.tsv()[2]) %>%
    mutate(sample_id = sub("-", ".", sample_id)) %>%
    mutate(group = paste(cell_type, treatment, sep = "."))
pd
```

<!-- Cell types A and B have been assay in a stimulted and non-stimulated -->
<!-- (none) conditions. There are 4 replicates in each of the 4 groups. -->


`r msmbstyle::question_begin()`
Before performing log-transformation of the data, check if there are
any 0-expression values. Indeed, these would be converted to `-Inf`
after log-transformation. Are there any 0-values in the data. If so,
how many are there.
`r msmbstyle::question_end()`


```{r, include = FALSE}
table(e == 0)
```

`r msmbstyle::question_begin()`
Add 1 to all expression values, then log-2 transform the expression
data. Visualise the distributions of the expression values in each
samples before and after transformation.
`r msmbstyle::question_end()`


```{r, fig.cap = "Expression values in cell types A and B without (black and gree) and with stimulation (red and blue).", warning = FALSE, include = FALSE}
boxplot(e, col = as.factor(pd$group), main = "Raw values")
e <- log2(e + 1)
boxplot(e, col = as.factor(pd$group), main = "Log-transformed")
```

Continue with log-transformed data, without preforming any normalisation.

`r msmbstyle::question_begin()`

Compute, for each gene, the number of 0-expression values and
visualise these as a table showing the number of genes with 0, 1, 2,
... zero values.

`r msmbstyle::question_end()`

```{r, include = FALSE}
table(row_zero <- apply(e, 1, function(x) sum(x == 0)))
```

`r msmbstyle::question_begin()`
Visualise and interpret the experiment design (cell type and
treatment) using a principal component analyis.
`r msmbstyle::question_end()`

```{r, fig.cap = "PCA of *kem2* samples.", include = FALSE}
library("factoextra")
pca_samples  <- prcomp(t(e), scale = TRUE, center = TRUE)
fviz_pca_ind(pca_samples, habillage = pd$group)
```
<!-- The PCA shows that samples group by treatment along the second PC -->
<!-- (22.3% of variance explained) and that close to 50% of the variance is -->
<!-- explained by variability along all samples. -->


`r msmbstyle::question_begin()`
Use a t test to identify the genes that are differentially expressed
between the stimulated and non-stimulated samples of cell type
A. Visualise the results on a volcano plot.
`r msmbstyle::question_end()`

<!-- We start by selecting the relevant samples and create a new expression -->
<!-- matrix matching cell type A.  -->
```{r, include = FALSE}
cell_type_a <- pd %>% filter(cell_type == "A")
cell_type_a

e_a <- e[, cell_type_a[[1]]]
e_a[1:10, 1:4]
```

<!-- Below, we perform a t test on each of the `r nrow(e_a)` genes. -->

```{r, fig.cap = "Histogram of p-values, showing an enrichment of small values.", include = FALSE}
my_t_test <- function(x)
    t.test(x[1:4], x[5:8])$p.value

res <- data.frame(row.names = rownames(e))
res$pv <- apply(e_a, 1, my_t_test)
hist(res$pv)
```

```{r, include = FALSE}
res$adjp <- p.adjust(res$pv, method = "BH")

my_lfc <- function(x)
    mean(x[1:4]) - mean(x[5:8])
res$lfc <- apply(e_a, 1, my_lfc)

res %>%
    arrange(adjp) %>%
    head()
```

```{r, fig.cap = "Volcano plot illustrating how stimulation down-regulates the expression of numerous genes.", include = FALSE}
with(res, plot(lfc, -log10(adjp)))
abline(v = c(-1, 1))
abline(h = -log10(0.001))
```

`r msmbstyle::question_begin()`
Considering that a gene is called differentially expressed if it has
an absolute log2 fold-change > 1 and an adjusted p-value < 0.001, how
many differentially genes are there?
`r msmbstyle::question_end()`


```{r, include = FALSE}
res %>%
    filter(adjp < 0.001, abs(lfc) > 1) %>%
    nrow
```

`r msmbstyle::question_begin()`
Visualise the expression of these differentially expressed genes in
all sampes using a heatmap and interpret the figure.
`r msmbstyle::question_end()`


```{r, fig.cap = "Heatmap of the 73 differentially expressed genes.", include = FALSE}
k_de <- res %>%
    rownames_to_column() %>%
    filter(adjp < 0.001, abs(lfc) > 1) %>%
    dplyr::select(rowname) %>%
    unlist()

heatmap(e[k_de, ])
```

```{r, message = FALSE, fig.cap = "Annotated heatmap of the 73 differentially expressed genes.", include = FALSE}
library("ComplexHeatmap")
col_annot <- HeatmapAnnotation(cell_type = pd$cell_type, treatment = pd$treatment)
Heatmap(e[k_de, ], top_annotation = col_annot)
```

<!-- We see that most genes that are differentially expressed in cell type -->
<!-- A are show a consistent pattern in cell type B, except for the botton -->
<!-- 5 genes, that show a low expression accross both conditions. -->


`r msmbstyle::question_begin()`
Select the 3 genes with the smallest p-values and visualise their
expression in all samples (cell types A and B, both stimulated and
un-stimulated) using boxplots.
`r msmbstyle::question_end()`



```{r, fig.cap = "Expression of the three most differentially expressed genes.", include = FALSE}
k_3 <- res %>%
    rownames_to_column() %>%
    arrange(adjp) %>%
    head(3) %>%
    dplyr::select(rowname) %>%
    unlist()

e[k_3, ] %>%
    as.data.frame() %>%
    rownames_to_column() %>%
    pivot_longer(names_to = "sample_id",
                 values_to = "expression",
                 -rowname) %>%
    full_join(pd) %>%
    ggplot(aes(x = treatment, y = expression)) +
    geom_boxplot(outlier.size = -1) +
    geom_jitter() +
    facet_grid(rowname ~ cell_type)
```


`r msmbstyle::question_begin()`
Perform a k-means clustering of all genes and samples using k = 5 and
visualise these clusters on a PCA plot. Interprete the figure.
`r msmbstyle::question_end()`


```{r, fig.cap = "PCA of the *kem2* genes displaying the 5 arbitrary clusters.", include = FALSE}
set.seed(1)
res$cl <- kmeans(e, centers = 5, nstart = 10)$cl
pca <- prcomp(e, scale = TRUE, center = TRUE)
fviz_pca_ind(pca, habillage = res$cl, geom = "point")
```

<!-- The clusters represent the diversity of gene expression along -->
<!-- PC1. Cluster 3, on the left of the figure, seems to be the most -->
<!-- variable and different, while to others represent a more continuous -->
<!-- gradient along of gene expression variation. -->


`r msmbstyle::question_begin()`
Repeat the volcano plot above, colouring the genes based on their
respective clusters. Refine your interpretation of the gene-level PCA
above.
`r msmbstyle::question_end()`



```{r, fig.cap = "Volcano plot displaying the five clusters", include = FALSE}
with(res, plot(lfc, -log10(adjp), col = res$cl))
legend("bottomleft", legend = 1:5, pch = 1, col = 1:5)
```

<!-- We see that the most diverse cluster corresponds roughly to the genes -->
<!-- that show significant down regulation upon stimulation. -->

