---
title: "Bioinformatics"
author: "Laurent Gatto"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
knit: bookdown::preview_chapter
description: "Course material for the Bbioinformatics (WSBIM1322) course at UCLouvain."
output:
  msmbstyle::msmb_html_book:
    toc: TRUE
    toc_depth: 1
    split_by: chapter
    split_bib: no
    css: style.css
link-citations: yes
bibliography: [refs.bib, packages.bib]
---

# Preamble {-}

The [WSBIM1322](https://uclouvain.be/cours-2019-wsbim1322.html) course
is teaches the basics of statistical data analysis applied to high
throughput biology. It is aimed at biology and biomedical students
that are already familiar with the R langauge (see the pre-requisits
section below). The students will familiarise themselves with
statitical learning concepts such as unsupervised and supervised
learning, hypothesis testing, and extend their understanding and
practive in R data structures and programming and the Bioconductor
project.

The course will be followed by *Omics data analysis*
([WSBIM2122](https://github.com/UCLouvain-CBIO/WSBIM2122)).


## Motivation {-}

Today, it is difficult to overestimate the very broad importance and
impact of *data*. Given the abundance of data around us, and the
sophistication of tools for their analysis and interpretation that are
readily available, data has become a tool of profound social
change. Resarch in general, and biomedical research in particular, is
at the centre of this evolution. And while bioinformatics has been
playing a central role in bio-medical research for many years now,
bioinformatics skills aren't well integrated in life science
curricula, limiting students in their career prospects and research
horizon [@WilsonSayres:2018]. It is important for young researchers to
acquire quantitative, computational and data skills to address the
challenges that lie ahead[^motiv1].

[^motiv1]: See also the [motivations behind the introductory
course](https://uclouvain-cbio.github.io/WSBIM1207/#motivation), which
still hold


This course will focus on the application of data analysis methods and
algorithms, and the interpretation of their outputs.  We will be using
the [R](https://www.R-project.org/) language and environment [@R] and
the [RStudio integrated development
environment](https://www.rstudio.com/products/RStudio/) to acquire
these data skills. Other interactive language such as
[Python](https://www.R-project.org/) and the interactive [jupyer
notebooks](https://jupyter.org/) would also have been a good fit. One
motivation of this choice is the availability of numerous
R/[Bionductor](https://www.bioconductor.org/) packages [@Huber:2015]
for the analysis of high throughput biology data.

## References and credits {-}

References are provided throughout the course. Several stand out
however, as they cover large parts of the material or provide
complementary resources.

- **Modern Statistical for Modern Biology**, by Susan Holmes and
  Wolfgang Huber [@MSMB]. A free online version of the book is
  available [here](https://www-huber.embl.de/msmb/).

- **An Introduction to Statistical Learning with Applications in R**
  by Gareth James, Daniela Witten, Trevor Hastie and Robert Tibshirani
  [@ISLR]. A free pdf of the book is available
  [here](http://faculty.marshall.usc.edu/gareth-james/ISL/index.html).

This course is being tought by Prof Laurent Gatto with invaluable
assistance from Dr Axelle Loriot at the Faculty of Pharmacy and
Biomedical Sciences (FASB) at the UCLouvain, Belgium.

## Pre-requisites {-}

Students taking this course should be familiar with data analysis and
visualisation in R. A formal pre-requisite for students taking the
class is the introductory course
[WSBIM1207](https://UCLouvain-CBIO.github.io/WSBIM1207). The first
chapter provides a refresher of the R skills needed for the rest of
the course.

Software requirements are documented in the *Setup* section below.

## About this course material {-}

This material is written in R markdown [@R-rmarkdown] and compiled as a
book using `knitr` [@R-knitr] `bookdown` [@R-bookdown]. The source
code is publicly available in a Github repository
[https://github.com/uclouvain-cbio/WSBIM1322](https://github.com/uclouvain-cbio/WSBIM1322)
and the compiled material can be read at http://bit.ly/WSBIM1322.

Contributions to this material are welcome. The best way to contribute
or contact the maintainers is by means of pull requests and
[issues](https://github.com/uclouvain-cbio/WSBIM1322/issues). Please
familiarise yourself with the [code of
conduct](https://github.com/UCLouvain-CBIO/WSBIM1322/blob/master/CONDUCT.md). By
participating in this project you agree to abide by its terms.

## Citation {-}

If you use this course, please cite it as

> Laurent Gatto. *UCLouvain-CBIO/WSBIM1322: Bioinformatics.*
> https://github.com/UCLouvain-CBIO/WSBIM1322.



## License {-}

This material is licensed under the [Creative Commons
Attribution-ShareAlike 4.0
License](https://creativecommons.org/licenses/by-sa/4.0/).


## Setup {-}

We will be using the [R environment for statistical
computing](https://www.r-project.org/) as main data science language.
We will also use the
[RStudio](https://www.rstudio.com/products/RStudio/) interface to
interact with R and write scripts and reports. Both R and RStudio are
easy to install and works on all major operating systems.

Once R and RStudio are installed, a set of packages will need to be
installed. See section \@ref(sec:setup2) for details.

The `rWSBIM1322` package provides some pre-formatted data used in this
course. It can be installed with

```{r, eval = FALSE}
BiocManager::install("UCLouvain-CBIO/rWSBIM1322")
```

and then loaded with

```{r rwsbim1322, message = FALSE, warning = FALSE}
library("rWSBIM1322")
```

To build this book, you'll need `bookdown` [@R-bookdown] and a
fork[^msmbfork] of [`msmbstyle`
style](https://github.com/grimbough/msmbstyle/) [@R-msmbstyle].

[^msmbfork]: https://github.com/lgatto/msmbstyle

```{r combilebook1, eval=FALSE}
install.packages("bookdown")
devtools::install_github("lgatto/msmbstyle")
```

In the course's work directory, simply type

```{r combilebook2, eval=FALSE}
bookdown::render_book(".")
```
